### 2018-02-01 submission/manuscript.tex
run spell-check

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of CRC. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma stages of CRC. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma stages of CRC. Next, using Random Forest models, we analyzed
whether the full community or only the combined top 5 increased and top
5 decreased RR genera resulted in better model classification, based on
the area under the curve (AUC). Finally, we also examined at what effect
and sample size the studies used were powered for and the sample size
needed to get to the traditionaly accepted 80\% power. Our results from
these analyses suggests that the bacterial community changes as disease
severity worsens, that specific members are important for disease
classification, and that many of the studies are underpowered for
assessing small effect sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of CRC:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overal median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

In stool, the most common genera in the top 10 most imporant variables,
in the full community models using the first approach, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
6A \& B{]}. Regardless of sample type, mouth-associated genera were
present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

@@

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies acheived 80\% power to detect a 30\% or greater difference
between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 7C{]}

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated micobe, and the results from this study suggest that
regardless of sample type, mouth-associated genera are commonly
associated with an increased RR for both adenoma and carcinoma stage of
disease.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for \(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.78) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.

@@

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples with control and CRC samples from different indivdiuals
were classified as unmatched while samples that belonged to the same
individual were classified as matched. Studies with matched data
included Burns, Dejea, Geng, and Lu while those with unmatched data were
from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
became unmatched when a corresponding matched sample did not make it
through sequence processing. All samples, from both tissue sample types,
were analyzed together for the linear mixed-effect models with samples
from the same individual corrected for. For all other analysis, not
mentioned herein, matched and unmatched samples were analyzed separately
using the statistical approaches mentioned in the Statistical Analysis
section.

>>>>

Using both feces (n = 1737) and colon tissues (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
and scope of the previous meta-analysis to investigate whether the
bacterial community or specific members are more important risk factors
for both adenoma and carcinoma stages of CRC. To accomplish this we
first assessed whether bacterial diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma stages of CRC. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma stages of CRC. Next, using Random Forest models, we analyzed
whether the full community or only the combined top 5 increased and top
5 decreased RR genera resulted in better model classification, based on
the area under the curve (AUC). Finally, we also examined at what effect
and sample size the studies used were powered for and the sample size
needed to get to the traditionally accepted 80\% power. Our results from
these analyses suggests that the bacterial community changes as disease
severity worsens, that specific members are important for disease
classification, and that many of the studies are underpowered for
assessing small effect sizes.

@@

\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
of CRC:}} To assess differences in broad scale community metrics as
disease severity worsens Operational Taxonomic Unit (OTU) richness,
evenness, and Shannon diversity measurements were power transformed and
Z-score normalized. These metrics are commonly used to assess the total
number of OTUs, the equality of their abundance, and the overall
diversity, respectively. Using linear mixed-effect models to control for
study and variable region we assessed whether OTU richness, evenness, or
Shannon diversity changed in a step-wise manner with disease severity.
In stool, there was a significant decrease in both evenness and Shannon
diversity as disease severity moved from control to adenoma to carcinoma
(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
tested whether the detectable differences in community significantly
increased in RR of having an adenoma or carcinoma. For fecal samples, a
decrease versus the overall median in evenness resulted in a
significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
0.032) while a decrease versus the overall median in Shannon diversity
only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
PERMANOVA, it was also possible to identify significant bacterial
community changes, in specific studies, for both carcinoma-associated
and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.

@@

In stool, the most common genera in the top 10 most important variables,
in the full community models using the first approach, were
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
6A \& B{]}. Regardless of sample type, mouth-associated genera were
present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

@@

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies achieved 80\% power to detect a 30\% or greater difference
between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 7C{]}

@@

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated microbe, and the results from this study suggest that
regardless of sample type, mouth-associated genera are commonly
associated with an increased RR for both adenoma and carcinoma stage of
disease.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for \(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hyper-variable
region used. Relative risk was analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.78) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.

@@

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples with control and CRC samples from different individuals
were classified as unmatched while samples that belonged to the same
individual were classified as matched. Studies with matched data
included Burns, Dejea, Geng, and Lu while those with unmatched data were
from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
became unmatched when a corresponding matched sample did not make it
through sequence processing. All samples, from both tissue sample types,
were analyzed together for the linear mixed-effect models with samples
from the same individual corrected for. For all other analysis, not
mentioned herein, matched and unmatched samples were analyzed separately
using the statistical approaches mentioned in the Statistical Analysis
section.
